---
title: "Getting Started with BJM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with BJM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

BJM fits a *backward joint model* of multivariate longitudinal outcomes and
time-to-event data, and uses it to make **dynamic predictions**: given a
patient's longitudinal history up to some `prediction_time`, predict their
risk of an event (and, if desired, their future biomarker values)
conditional on survival to that point.

Fitting and prediction is a four-step pipeline:

1. `survivalSub()` fits the marginal survival sub-model (and, optionally, a
   competing-risks sub-model).
2. `longitudinalSub()` fits the longitudinal sub-model(s), one per
   biomarker.
3. `predictRisk()` combines the two fits to predict a patient's event
   risk over a future horizon.
4. `predictLongitudinal()` additionally predicts a future value of one or
   more specific biomarkers.

This vignette walks through all four steps on the `pbc3` data set that
ships with the package, then covers two practical questions that come up
once the mechanics are working: how to choose the `bandcount1`/
`bandcount2`/`bandcount3` numerical-integration tuning parameters, and what
happens when the pipeline functions are called with malformed inputs.

```{r setup}
library(BJM)
data(pbc3)
```

`pbc3` is in long format: one row per longitudinal measurement, with
baseline covariates (`age`, `sex`, ...) repeated on every row for a given
patient, and event-time variables (`years`, `status3`, ...) that are also
constant within patient.

## Step 1: Fit the survival sub-model with `survivalSub()`

`survivalSub()` needs one row per patient, so we first drop the repeated
longitudinal rows. `form_marginal_surv` is a standard `survival::Surv()`
formula for the overall event. `form_conditional_cr` is optional: supply it
when there is more than one competing event type and you want to predict
event-specific risk (here, `status4` distinguishes the two causes among
patients who had *some* event).

```{r survival-sub}
data_survival_fitting <- pbc3[!duplicated(pbc3$id), ]

survival_fit_all <- survivalSub(
  data_survival_fitting,
  form_marginal_surv = Surv(years, status3) ~ age + sex,
  form_conditional_cr = status4 ~ years + age + sex
)

survival_fit_all
```

## Step 2: Fit the longitudinal sub-models with `longitudinalSub()`

Each biomarker gets its own fixed-effects formula and random-effects
formula, supplied as same-length, same-order lists. `data_fit_all` is a
matching list of fitting data sets (one per biomarker) — or a single
`data.frame` that is reused for every biomarker, if all biomarkers are fit
on the same data. Here we fit two biomarkers, `serBilir` and `albumin`, on
the complete-case subset.

```{r longitudinal-sub}
long_sub_fixed <- list(
  "serBilir" = serBilir ~ year + age + sex + (years) + (years) * year,
  "albumin"  = albumin  ~ year + age + sex + (years) + (years) * year
)
long_sub_random <- list(
  "serBilir" = ~ year | id,
  "albumin"  = ~ year | id
)

data_fit_all <- list(pbc3[pbc3$status3 == 1, ], pbc3[pbc3$status3 == 1, ])

long_fit_all <- longitudinalSub(data_fit_all, long_sub_fixed, long_sub_random)

long_fit_all
```

## Optional: imputing interrupted follow-up before fitting

`longitudinalSub()`, as used above, does complete-case analysis: for each
biomarker it drops rows missing that biomarker's own covariates, and then
keeps only the subjects who have at least one non-missing observation of
*every* biomarker. With interrupted/irregular follow-up this throws away
real information -- a subject who has `serBilir` at every visit but never
had `albumin` drawn is dropped from *both* models, not just the `albumin`
one.

`imputeLongitudinal()` is an optional preprocessing step that fills these
gaps before `longitudinalSub()` runs, assuming the missingness is at random
(MAR) given the observed covariates and biomarkers. Two backends are
available, selected with `method`:

- `method = "miwae"` (default) fits a MIWAE (Missing data
  Importance-Weighted AutoEncoder; Mattei & Frellsen, 2019) jointly across
  the supplied biomarkers and draws plausible completions via
  self-normalized importance resampling.
- `method = "diffusion"` fits a conditional denoising diffusion
  probabilistic model (DDPM; Ho, Jain & Abbeel, 2020) instead, trained with
  a masked score-matching loss, and imputes with a RePaint-style (Lugmayr
  et al., 2022) reverse-diffusion sampler that reproduces every observed
  value exactly and only draws the missing ones.

Both share the same call shape and return shape, so switching between them
is a one-argument change; the completed data can be passed straight to
`longitudinalSub()` in place of the original `data_fit_all` either way. It
requires the optional `torch` package (listed in `Suggests`, not
`Imports`), so the rest of this section only runs if `torch` is installed.

```{r impute-check}
torch_available <- requireNamespace("torch", quietly = TRUE) &&
  isTRUE(tryCatch(torch::torch_is_installed(), error = function(e) FALSE))
torch_available
```

To see it in action, artificially delete 12% of the `serBilir` and
`albumin` measurements in the fitting data at random -- satisfying MAR by
construction, and mimicking a subject whose visit happened but one
biomarker wasn't drawn that day:

```{r impute-mask, eval = torch_available}
set.seed(1)
data_fit_masked <- data_fit_all[[1]]
n <- nrow(data_fit_masked)
data_fit_masked$serBilir[sample.int(n, floor(0.12 * n))] <- NA
data_fit_masked$albumin[sample.int(n, floor(0.12 * n))] <- NA
```

```{r impute-run, eval = torch_available}
imputed <- imputeLongitudinal(
  data_fit_masked, long_sub_fixed, long_sub_random,
  time_variable = "year", epochs = 80, seed = 1
)
```

`diagnostics` reports, per biomarker, the observed mean/sd alongside the
deep-generative and classical-baseline imputed mean/sd -- useful for
judging whether the deep model's completions look reasonable before
trusting them on your own data, since a deep generative model needs enough
data to fit reliably and is not automatically the better choice at every
sample size:

```{r impute-diagnostics, eval = torch_available}
imputed$diagnostics$serBilir
```

Feeding the completed data into `longitudinalSub()` retains at least as
many subjects as running it on the masked data directly, because a subject
who was excluded only for missing one biomarker at one visit is no longer
excluded once that cell is filled in:

```{r impute-subjects, eval = torch_available}
count_retained <- function(data_fit_all, long_sub_fixed, id = "id") {
  if (is.data.frame(data_fit_all)) {
    data_fit_all <- rep(list(data_fit_all), length(long_sub_fixed))
  }
  ids_by_marker <- lapply(seq_along(long_sub_fixed), function(m) {
    d <- data_fit_all[[m]]
    complete <- d[rowSums(is.na(d[all.vars(long_sub_fixed[[m]])])) == 0, ]
    unique(complete[[id]])
  })
  Reduce(intersect, ids_by_marker)
}

length(count_retained(data_fit_masked, long_sub_fixed))
length(count_retained(imputed$data_fit_all, long_sub_fixed))
```

```{r impute-fit, eval = torch_available}
long_fit_imputed <- longitudinalSub(imputed$data_fit_all, long_sub_fixed, long_sub_random)
```

### A diffusion-based alternative

The same masked data can be completed with `method = "diffusion"` instead
-- same arguments, a different generative model underneath:

```{r impute-run-diffusion, eval = torch_available}
imputed_diffusion <- imputeLongitudinal(
  data_fit_masked, long_sub_fixed, long_sub_random,
  time_variable = "year", method = "diffusion",
  epochs = 60, diffusion_steps = 40, seed = 1
)
imputed_diffusion$diagnostics$serBilir
```

```{r impute-fit-diffusion, eval = torch_available}
long_fit_imputed_diffusion <- longitudinalSub(imputed_diffusion$data_fit_all,
                                               long_sub_fixed, long_sub_random)
length(count_retained(imputed_diffusion$data_fit_all, long_sub_fixed))
```

Neither backend is automatically the better choice: compare the
`diagnostics` above against the observed mean/sd, and, if in doubt, compare
downstream fixed-effect estimates from `long_fit_imputed` and
`long_fit_imputed_diffusion` against `long_fit_all` (the original
complete-case fit) before relying on either for inference.

### Propagating imputation uncertainty with `poolLongitudinalSub()`

`imputed` above used the default `impute = "single"`: `data_fit_all` is
filled with the *mean* of the completions drawn from the generative model,
giving one completed dataset. That is convenient, but it understates
uncertainty -- fitting `longitudinalSub()` on a single completed dataset
treats the imputed values as if they were observed data, so the resulting
standard errors do not reflect not actually knowing them.

`impute = "multiple"` instead returns `n_imputations` separately completed
datasets in `data_fit_all_list`, without averaging them together. Fitting
`longitudinalSub()` once per completion and combining the fixed-effect
estimates with Rubin's rules (Rubin, 1987) -- implemented here by
`poolLongitudinalSub()` -- propagates the disagreement between completions
into the final standard errors, instead of discarding it:

```{r impute-run-multiple, eval = torch_available}
imputed_multiple <- imputeLongitudinal(
  data_fit_masked, long_sub_fixed, long_sub_random,
  time_variable = "year", n_imputations = 5, impute = "multiple",
  epochs = 80, seed = 1
)
length(imputed_multiple$data_fit_all_list)
```

```{r impute-pool, eval = torch_available}
long_fit_all_list <- lapply(imputed_multiple$data_fit_all_list, function(d) {
  longitudinalSub(d, long_sub_fixed, long_sub_random)
})
pooled <- poolLongitudinalSub(long_fit_all_list)
pooled
```

The printed table's `FMI` (fraction of missing information) column reports,
per coefficient, how much precision the missingness cost relative to having
observed everything -- a coefficient with a high FMI is one where the
choice to impute (and how) matters more for that specific estimate, even if
the biomarker's overall missingness rate looks modest. `pooled$pooled` is a
named list (one `data.frame` per biomarker) with the pooled `estimate`,
`std_error`, `df`, `statistic`, `p_value`, `riv`, and `fmi` for programmatic
use.

## Step 3: Predict event risk with `predictRisk()`

To predict for a specific patient, build a `data_predict_all` list (one
`data.frame` per biomarker, mirroring `data_fit_all`) containing only that
patient's measurements *up to* `prediction_time` — later measurements
would not be available yet in a real prediction setting. Below we predict,
for patient 2, the risk of each event type within one year of `year = 5`.

`survival_variable_all`/`survival_trans_function` describe how the raw
event-time variable is transformed for the integration grid; see
`?predictRisk` for details.

```{r dynamic-prediction}
survival_variable_all <- list("Tyears1", "Tyears2", "Tyears3", "Tyears4")
survival_trans_function <- list(
  fun1 = function(x) abs(x - 1),
  fun2 = function(x) abs(x - 3),
  fun3 = function(x) abs(x - 5),
  fun4 = function(x) abs(x - 7)
)

data_raw_predict <- pbc3[pbc3$id == 2 & pbc3$year <= 5, ]
data_predict_all <- list(data_raw_predict, data_raw_predict)

risk <- predictRisk(
  data_predict_all, long_fit_all, survival_fit_all,
  prediction_time = 5, horizon = 1, time_variable = "year",
  survival_variable_all, survival_trans_function,
  bandcount1 = 10, bandcount2 = 20
)

risk
```

`survival_variable_all`/`survival_trans_function` almost always follow the
same convention shown above: variables named `"Tyears1"`, `"Tyears2"`, ...,
each the absolute distance from a fixed cut point. `survivalTrans()` builds
exactly that pair from a plain vector of cut points, so you do not have to
hand-write two matching parallel lists:

```{r survival-trans-helper}
trans <- survivalTrans(c(1, 3, 5, 7))
identical(trans$survival_variable_all, survival_variable_all)
trans$survival_trans_function[[1]](2)
```

`risk_prob_1` and `risk_prob_2` are the predicted probabilities of
experiencing each of the two competing event types within the one-year
horizon, conditional on the patient's longitudinal history and survival to
`prediction_time`. (`risk_prob_2` is `NULL` whenever `survival_fit_all` was
fit without `form_conditional_cr`.)

## Step 4: Predict a future biomarker value with `predictLongitudinal()`

`predictLongitudinal()` answers a different question: not *whether* an
event happens, but what a specific biomarker's value is likely to be at
`prediction_time + horizon`, conditional on survival. `bio_i` selects the
biomarker(s) by position in `long_fit_all` (`1` = `serBilir` here); naming
exactly one biomarker returns a single result, while naming more than one
(or leaving `bio_i = NULL` for every biomarker) returns a named list of
results instead, computing the shared, bio_i-independent part of the
pipeline only once and reusing it across every biomarker.

```{r dynamic-prediction-bio}
bio_pred <- predictLongitudinal(
  bio_i = 1, data_predict_all, long_fit_all, survival_fit_all,
  prediction_time = 5, horizon = 1, time_variable = "year",
  survival_variable_all, survival_trans_function,
  bandcount2 = 20, bandcount3 = 50
)

bio_pred$Y_predict
```

`Y_predict` is the MAP (most likely) predicted value; `Y_density`/`Y_all`
give the full predicted density over a grid of candidate values, which is
what `predictPlot()` visualizes (see `?predictPlot`).

## Choosing `bandcount1`, `bandcount2`, `bandcount3`

`predictRisk()` and `predictLongitudinal()` estimate their outputs
by numerical integration over patient-specific time and biomarker grids.
The `bandcount*` arguments control how fine those grids are:

- **`bandcount1`** (`predictRisk()` only) is the number of grid
  points spanning the prediction window itself, from `prediction_time` to
  `prediction_time + horizon`. This is the numerator of the risk
  probability.
- **`bandcount2`** is the number of grid points spanning
  `[prediction_time, upper_bound]`, where `upper_bound` is set internally
  to twice the longest observed survival/censoring time among at-risk
  patients. This approximates "integrating out to infinity" for the
  denominator that normalizes the risk probability. A wider follow-up
  range needs a larger `bandcount2` to keep the grid spacing comparable.
- **`bandcount3`** (`predictLongitudinal()` only) is the number of points
  in the candidate-biomarker-value grid (`Y_all`) used to build the
  predicted density and locate its mode (`Y_predict`). It controls the
  resolution of the density curve, not a time integral.

Larger values give more accurate, smoother results at the cost of more
computation. There is no universal "correct" value because it depends on
how quickly the underlying hazard and biomarker trajectories change and on
the length of follow-up in your data — so the practical approach is a
**convergence check**: run the prediction once with the defaults, once with
every `bandcount*` doubled, and confirm the results barely move.

```{r bandcount-convergence}
risk_default <- predictRisk(
  data_predict_all, long_fit_all, survival_fit_all,
  prediction_time = 5, horizon = 1, time_variable = "year",
  survival_variable_all, survival_trans_function,
  bandcount1 = 10, bandcount2 = 20
)

risk_doubled <- predictRisk(
  data_predict_all, long_fit_all, survival_fit_all,
  prediction_time = 5, horizon = 1, time_variable = "year",
  survival_variable_all, survival_trans_function,
  bandcount1 = 20, bandcount2 = 40
)

abs(risk_default$risk_prob_1 - risk_doubled$risk_prob_1)
abs(risk_default$risk_prob_2 - risk_doubled$risk_prob_2)
```

If doubling the `bandcount*` values changes the result by more than you can
tolerate, keep doubling until it doesn't; if it barely changes anything (as
above), the smaller, cheaper value is fine to use. The same check applies
to `bandcount3` for `predictLongitudinal()`.

## Friendly error messages

The pipeline functions validate their arguments before doing any model
fitting or numerical integration, so common mistakes fail fast with a
message that names the offending argument — instead of a cryptic error
from deep inside `model.matrix()` or list indexing. For example, passing a
single `data.frame` instead of a list to `predictRisk()`:

```{r validation-example, error = TRUE}
predictRisk(
  data_predict_all[[1]], long_fit_all, survival_fit_all,
  prediction_time = 5, horizon = 1, time_variable = "year",
  survival_variable_all, survival_trans_function,
  bandcount1 = 10, bandcount2 = 20
)
```

or a formula referencing a column that doesn't exist:

```{r validation-example-2, error = TRUE}
longitudinalSub(pbc3, serBilir ~ year + not_a_column, ~ year | id)
```

## Where to go next

- `?predictPlot` and `?riskPlot` visualize the outputs of
  `predictLongitudinal()`/`predictRisk()` for a single patient.
- `?cmtPlot` plots observed longitudinal trajectories stratified by
  eventual outcome, useful for checking whether a biomarker looks
  informative before fitting.
- `?spaghettiPlot` draws each subject's observed trajectory, optionally
  aligned at the event time, and `?cifPlot` plots the cumulative
  incidence of each (competing) event type.
- `?performancePlot` evaluates the risk predictions (time-dependent AUC
  and Brier score) across landmark times, ideally on held-out data, and
  `?calibrationPlot` compares the predicted with the observed risks.
- `?imputeLongitudinal` fills in missing longitudinal biomarker values due
  to interrupted follow-up before `longitudinalSub()` runs, instead of
  relying on complete-case analysis (see "Optional: imputing interrupted
  follow-up before fitting" above).
- `?poolLongitudinalSub` combines `longitudinalSub()` fits across the
  multiple completed datasets from `imputeLongitudinal(..., impute =
  "multiple")` with Rubin's rules, so imputation uncertainty is reflected
  in the reported standard errors (see "Propagating imputation uncertainty
  with `poolLongitudinalSub()`" above).
