---
title: "Nonlinear terms in longitudinal formulas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nonlinear terms in longitudinal formulas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`longitudinalSub()` fits each biomarker's fixed effects (`long_sub_fixed`)
and random effects (`long_sub_random`) as ordinary `nlme::lme()` formulas, so
any term `lme()` accepts -- polynomials, splines, factors, interactions,
log/sqrt transforms -- can appear on the right-hand side. The one thing that
matters for *correctness*, not just syntax, is whether a term's basis
depends on the data it is computed from:

- `poly(x, df, raw = TRUE)`, `I(x^2)`, `log(x)`, `sqrt(x)`, and their
  interactions compute the same value no matter what other rows are present.
  These are always safe, everywhere.
- `poly(x, df)` in its default **orthogonal** mode, `splines::ns()`/
  `splines::bs()`, and `factor()` instead compute their basis/contrasts from
  *whatever data they are given*. `longitudinalSub()` only caches and reuses
  the training-time basis for a **continuous** biomarker's `long_sub_fixed`
  formula. Everywhere else -- `long_sub_random`, or an **ordinal**
  biomarker's `long_sub_fixed` -- the design matrix is rebuilt from a single
  patient's rows at prediction time, so the basis can silently disagree with
  the one the model was fit with. `longitudinalSub()` emits a warning
  whenever one of these appears somewhere it isn't cached.

|  | continuous `long_sub_fixed` | ordinal `long_sub_fixed` | `long_sub_random` |
|---|---|---|---|
| `poly(x, df, raw = TRUE)`, `I(x^2)`, `log(x)`, `sqrt(x)`, interactions of these | safe | safe | safe |
| `poly(x, df)` (orthogonal, the default), `splines::ns()`/`splines::bs()`, `factor()` | safe (basis cached and reused) | warns | warns |

This vignette walks through both groups on `pbc3`, shows what the warning
looks like when a data-dependent term ends up somewhere unsafe, and finishes
with a spline term carried all the way through to `predictRisk()`.

```{r setup}
library(BJM)
data(pbc3)
data_fit <- pbc3[pbc3$status3 == 1, ]
```

## Terms that are always safe

### Explicit polynomials and transforms

`poly(..., raw = TRUE)` returns the raw powers `x, x^2, ..., x^df` instead of
an orthogonalized basis, so (unlike the default `poly()`) it computes the
same columns regardless of what other data happen to be present:

```{r poly-raw}
fit_poly_raw <- longitudinalSub(
  data_fit, serBilir ~ poly(year, 2, raw = TRUE) + age + sex, ~ year | id
)
```

`I(x^2)`, `log(x)`, and `sqrt(x)` are even simpler -- each is just an
ordinary arithmetic transform of one column, with nothing to disagree about:

```{r explicit-transforms}
fit_explicit <- longitudinalSub(
  data_fit, serBilir ~ log(year + 1) + sqrt(year) + age + sex, ~ year | id
)
```

### Interactions

Interactions of safe terms are themselves safe -- e.g. letting the
quadratic time trend differ by age:

```{r interaction}
fit_interaction <- longitudinalSub(
  data_fit, serBilir ~ year + I(year^2) * age + sex, ~ year | id
)
```

## Terms that are safe only in a continuous biomarker's `long_sub_fixed`

### Orthogonal `poly()`, splines, and `factor()`

Used in a continuous biomarker's `long_sub_fixed`, these are fine: the basis
`lme()` was fit with, and the factor levels observed in the full training
data, are cached on the returned `longitudinalSub.BJM` object and reused
both when estimating `Sigma_fit` and at prediction time -- never recomputed
from a smaller slice of data.

```{r safe-in-fixed}
fit_orthogonal_poly <- longitudinalSub(
  data_fit, serBilir ~ year + poly(age, 2) + factor(sex), ~ year | id
)

fit_ns <- longitudinalSub(
  data_fit, serBilir ~ splines::ns(year, df = 3) + age + sex, ~ year | id
)
```

## What goes wrong elsewhere

The same `splines::ns()` term in `long_sub_random` -- instead of
`long_sub_fixed` -- is not cached, because `long_sub_random`'s design matrix
is always rebuilt from whatever rows are at hand. `longitudinalSub()` warns
rather than silently risking a basis mismatch:

```{r unsafe-warning, warning = TRUE}
fit_ns_random <- longitudinalSub(
  data_fit, serBilir ~ year + age + sex, ~ splines::ns(year, df = 2) | id
)
```

The same warning fires for an **ordinal** biomarker's `long_sub_fixed`,
since ordinal fits (via `ordinal::clmm()`) rebuild their design matrix from
the patient-specific data at prediction time the same way `long_sub_random`
does. In both cases, switch to `poly(..., raw = TRUE)`, `I(x^2)`, `log()`,
or `sqrt()`.

## Precomputed time transforms with `survivalTrans()`

A transformed-time column built once, ahead of fitting, is just an ordinary
numeric column by the time `long_sub_fixed` sees it -- always safe,
including in `long_sub_random`:

```{r survival-trans}
trans <- survivalTrans(c(1, 3, 5, 7))
data_fit_trans <- data_fit
data_fit_trans$Tyears1 <- trans$survival_trans_function[[1]](data_fit_trans$year)

fit_trans <- longitudinalSub(
  data_fit_trans, serBilir ~ year + Tyears1 + age + sex, ~ year | id
)
```

## Putting it together: a spline term through to prediction

The cached-basis guarantee above means a spline fixed effect is not just
safe to *fit* -- predictions built on it are exactly as valid as with a
linear term. `years` has to appear in `long_sub_fixed` as well (the
convention every BJM formula follows, so the model can be evaluated at the
grid of remaining-survival-time values `predictRisk()` integrates over):

```{r end-to-end}
fit_spline <- longitudinalSub(
  data_fit, serBilir ~ splines::ns(year, df = 3) + age + sex + years, ~ year | id
)

data_survival_fitting <- pbc3[!duplicated(pbc3$id), ]
survival_fit <- survivalSub(
  data_survival_fitting,
  form_marginal_surv = Surv(years, status3) ~ age + sex,
  form_conditional_cr = NULL
)

data_predict <- pbc3[pbc3$id == 2 & pbc3$year <= 5, ]
risk <- predictRisk(
  data_predict, fit_spline, survival_fit,
  prediction_time = 5, horizon = 1, time_variable = "year",
  survival_variable_all = list(), survival_trans_function = list(),
  bandcount1 = 10, bandcount2 = 20
)
risk
```

See `vignette("BJM-intro", package = "BJM")` for the full four-step
pipeline, and `?longitudinalSub` for the complete list of nonlinear-term
examples shown there.
