---
title: "Interval-censored events (experimental)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

In many longitudinal studies the event is not seen when it happens. It is
found at a visit: a biopsy turns positive, a scan shows progression, a
cognitive test crosses a threshold. All that is known is that the event
happened after the last visit at which it was absent and before the visit
that detected it. The event time is **interval-censored**, known only to
lie in an interval $(L, R]$:

| Situation | $L$ | $R$ |
|---|---|---|
| Event detected at a visit | last negative visit | the detecting visit |
| Event already present at the first visit | 0 | first visit |
| No event by the last visit (right-censored) | last visit | `NA` |

Treating the detecting visit or the midpoint of the interval as the event
time is common, but it biases the backward joint model in a particular
way. The longitudinal sub-model $f(Y \mid T)$ has the event time $T$ as a
covariate, so an error in $T$ is measurement error in a covariate, and its
coefficients are attenuated. This article shows the experimental
interval-censoring support in BJM:

- `survivalSub()` fits the survival sub-model to `Surv(L, R, type = "interval2")`;
- `fitIntervalBJM()` fits the longitudinal sub-model by imputing $T$;
- `predictRisk()` and `predictLongitudinal()` account for the event being
  possibly already present, but not yet detected, at the prediction time;
- `performancePlot()` and `calibrationPlot()` evaluate the predictions.

```{r setup, message = FALSE}
library(BJM)
```

## Example data

The `pbc3` data have exact event times, so this article simulates a study
whose truth is known. Subjects are seen every 1 to 3 years, when a
biomarker `y` is measured and the event is checked for. The event time
follows a proportional hazards model with covariate `x`, and the biomarker
depends on the (unknown) event time `T` with coefficient −0.2.

```{r data}
simulate_study <- function(n, seed) {
  set.seed(seed)
  x <- rnorm(n)
  T <- (-log(runif(n)) / (0.05 * exp(0.5 * x)))^(1 / 1.5)
  # true_T is kept only to check the methods below; a real study does not have it
  surv <- data.frame(id = seq_len(n), x = x, L = NA_real_, R = NA_real_, true_T = T)
  long <- vector("list", n)
  for (i in seq_len(n)) {
    visits <- c(0, cumsum(runif(20, 1, 3)))
    visits <- visits[visits < runif(1, 4, 15)]
    if (T[i] > max(visits)) {                 # not detected: right-censored
      surv$L[i] <- max(visits)
      seen <- visits
    } else {                                  # detected at the next visit
      k <- findInterval(T[i], visits)
      surv$L[i] <- visits[k]
      surv$R[i] <- visits[k + 1]
      seen <- visits[visits <= T[i]]
    }
    b <- rnorm(2, 0, c(0.5, 0.1))
    long[[i]] <- data.frame(id = i, year = seen, x = x[i],
                            y = 1 + 0.3 * seen - 0.2 * T[i] + 0.4 * x[i] +
                              b[1] + b[2] * seen + rnorm(length(seen), 0, 0.3))
  }
  list(surv = surv, long = do.call(rbind, long))
}
study <- simulate_study(300, seed = 1)
head(study$surv)
head(study$long)
```

## The survival sub-model

`Surv(L, R, type = "interval2")` makes `survivalSub()` fit a proportional
hazards model to the intervals by maximum likelihood, in place of a Cox
model. The baseline hazard is a smooth Royston–Parmar spline by default
(`baseline = "spline"`, `df = 3`); `baseline = "piecewise"` gives a
piecewise-constant hazard. Because the exact event time is unknown, it is
not part of the formula: `event_time` names the column that the
longitudinal sub-model will use for it.

```{r survival}
survival_fit <- survivalSub(study$surv, Surv(L, R, type = "interval2") ~ x, NULL,
                            event_time = "Tev")
survival_fit
```

The printout reports how many events fell **before the first visit**.
Before the earliest visit there are no interval endpoints, so the baseline
hazard there is extrapolated rather than estimated. When such events are a
large share of all events (in simulations, about a third or more), results
depend noticeably on that extrapolation.

A `strata()` term gives each stratum its own baseline hazard, as in a
stratified Cox model.

## Fitting the longitudinal sub-model: `fitIntervalBJM()`

With right-censored data, the longitudinal sub-model is fit to the subjects
whose event was observed, using their event times. With interval censoring
no event time is known. `fitIntervalBJM()` fills it in by stochastic EM and
multiple imputation:

1. fit the survival sub-model to the intervals;
2. start every event time at its interval midpoint and fit
   `longitudinalSub()`;
3. draw each subject's $T$ from
   $f(T \mid L < T \le R, Y) \propto f(Y \mid T)\, f(T)$ on their
   interval, and refit, for `n_burnin` iterations;
4. keep iterating, saving `n_imputations` draws and their fits.

Step 3 uses the backward model's own structure: $f(Y \mid T)$ on a grid of
event times is exactly what dynamic prediction already computes.

```{r fit}
fit <- fitIntervalBJM(
  study$surv, study$long,
  form_marginal_surv = Surv(L, R, type = "interval2") ~ x,
  event_time = "Tev",
  long_sub_fixed = y ~ year + Tev + x,
  long_sub_random = ~ year | id,
  time_variable = "year",
  n_burnin = 5, n_imputations = 3, seed = 1)
fit$pooled$pooled[[1]][, c("estimate", "std_error", "fmi")]
```

`fit$pooled` combines the imputations with Rubin's rules
(`poolLongitudinalSub()`); `fmi` is the fraction of missing information,
here the information lost by not knowing $T$. `fit$trace` holds every iteration's
estimates, to check that the chain has settled:

```{r trace, fig.height = 3.5, fig.alt = "Coefficient of the event time in the longitudinal sub-model at each stochastic EM iteration; the values fluctuate in a narrow band around -0.20, with a dashed line marking the end of the burn-in."}
plot(fit$trace[, "Tev"], type = "b", xlab = "Iteration",
     ylab = "Coefficient of Tev", main = "Stochastic EM trace")
abline(v = 5.5, lty = 2)
```

For comparison, the midpoint fit and the fit with the true event times,
which a real study would not have:

```{r compare}
events <- study$surv[!is.na(study$surv$R), ]
with_T <- function(T_values) {
  d <- study$long[study$long$id %in% events$id, ]
  d$Tev <- T_values[match(d$id, events$id)]
  nlme::fixef(longitudinalSub(list(d), y ~ year + Tev + x, ~ year | id)$lfit[[1]])
}
round(rbind(truth = c(1, 0.3, -0.2, 0.4),
            `true event times` = with_T(events$true_T),
            midpoint = with_T((events$L + events$R) / 2),
            fitIntervalBJM = fit$pooled$pooled[[1]]$estimate), 3)
```

One data set says little. In simulations of this design (300 subjects,
many replicates), the midpoint fit's coefficient of `Tev` was attenuated by
about 4%, and `fitIntervalBJM()` removed this. With visits only every 2–6
years the midpoint bias grew to about 9%, and `fitIntervalBJM()` left about
1–2%.

## Dynamic prediction between visits

`fit$survival_fit_all` and `fit$long_fit_all` plug into the usual prediction
functions. One thing changes: an event is only detected at a visit, so at a
prediction time $s$ between visits, a patient is known to be event-free only
up to their last visit $V$, not up to $s$. `predictRisk()` therefore returns

- `risk_prob_1`: $P(s < T \le s + \text{horizon} \mid T > V, \text{history})$;
- `prob_undetected_1`: $P(V < T \le s \mid T > V, \text{history})$, the
  probability that the event has already happened but has not been detected yet.

```{r predict}
# six patients still followed at year 4 with no event detected by then
at_risk <- with(study$surv, id[(is.na(R) & L >= 4) | (!is.na(R) & R > 4)])[1:6]
history <- study$long[study$long$id %in% at_risk & study$long$year <= 4, ]
history$Tev <- NA          # unknown for a patient still being followed
tapply(history$year, history$id, max)   # each patient's last visit V
risk <- predictRisk(list(history), fit$long_fit_all, fit$survival_fit_all,
                    prediction_time = 4, horizon = 2, time_variable = "year",
                    survival_variable_all = NULL, survival_trans_function = NULL,
                    bandcount1 = 20, bandcount2 = 40)
risk
```

Here the simulation's true event times show what the model could not
see. A high `Already, undetected` probability goes with an event that had
in fact already happened by year 4:

```{r predict-truth}
round(study$surv$true_T[match(names(risk$risk_prob_1), study$surv$id)], 2)
```

Predicting at a patient's last visit ($s = V$) leaves nothing undetected and
gives the usual conditioning on $T > s$. `simulateTrajectory()` currently
requires this, since the event could otherwise already have happened.

## Competing risks

With competing risks, the event type is taken as known once the event is
detected. Pass the event-type model to `fitIntervalBJM()`. It draws each
$T$ from $f(Y \mid T, D)\, P(D \mid T)\, f(T)$ and refits the event-type
model, which uses $T$, on every draw. `predictRisk()` then reports both
risks and both undetected probabilities:

```{r competing, eval = FALSE}
fit_cr <- fitIntervalBJM(
  surv, long, Surv(L, R, type = "interval2") ~ x, "Tev",
  long_sub_fixed = y ~ year + Tev + type + x,
  long_sub_random = ~ year | id, time_variable = "year",
  form_conditional_cr = type ~ Tev + x)
```

The event type (`type`, 0/1) must be in both data sets, as for
right-censored competing risks.

## Evaluating the predictions

`performancePlot()` and `calibrationPlot()` take evaluation data with each
subject's interval columns (here `L` and `R`) instead of a survival time and
status. A subject whose interval straddles the start or end of the
prediction window has an uncertain status. Following Yang et al. (2026),
the default (`interval_method = "model"`) counts such a subject as a case
or a control with the probability the fitted model gives their event time
within their own interval.

```{r performance, fig.height = 3.5, fig.alt = "Time-dependent AUC and Brier score at landmark times 2, 4 and 6, model-based."}
evaluation <- merge(study$long, study$surv[, c("id", "L", "R")], by = "id")
performancePlot(evaluation, fit$long_fit_all, fit$survival_fit_all,
                prediction_time = c(2, 4, 6), horizon = 2, time_variable = "year",
                survival_variable_all = NULL, survival_trans_function = NULL,
                bandcount1 = 20, bandcount2 = 40)
```

`interval_method = "ipcw"` uses only the subjects whose status is certain,
reweighted, and does not rely on the model. In simulations with visits
every 1–3 years, though, only about 5 of 60 cases had a certain status. The
IPCW Brier score came out at about 0.03 instead of the true 0.13, and the
IPCW AUC was about three times as variable. The model-based measures stayed
within about 0.003 of the truth, even with a misspecified survival
sub-model. Because the model grades itself, a miscalibrated model's Brier
score comes out slightly too good, so check calibration separately with
`calibrationPlot()`, which does not use the model.

In `calibrationPlot()`, each risk group's observed risk is estimated
without the model. A Turnbull-type nonparametric estimate is computed from
the intervals and left-truncated at each subject's last visit, so it
targets the same quantity `predictRisk()` predicts:

```{r calibration, fig.height = 4, fig.alt = "Calibration plot: mean predicted risk against the nonparametric observed risk in four risk groups, at landmarks 2 and 4."}
calibrationPlot(evaluation, fit$long_fit_all, fit$survival_fit_all,
                prediction_time = c(2, 4), horizon = 2, time_variable = "year",
                survival_variable_all = NULL, survival_trans_function = NULL,
                n_groups = 4, bandcount1 = 20, bandcount2 = 40)
```

In simulations with visits every 1–3 years and a 2-year window, this
observed risk was unbiased for every risk group. It also exposed a
deliberately miscalibrated model, where an observed risk computed from the
model's own posterior followed the miscalibrated predictions instead. But
with event times known only up to their intervals it is noisy: with groups
of about 60 subjects it was off by 0.08–0.16 on average, so a group can sit
well off the diagonal by chance, as at landmark 4 here. It also needs the
prediction window to be at least about as long as the gaps between visits.
With visits every 2–6 years and a 2-year window, the window usually lies
inside a subject's interval, where the timing is not identified, and the
observed risk of the higher-risk groups came out up to 0.27 too high. As
everywhere in BJM, evaluating on the fitting data gives an optimistic
(apparent) performance. Use held-out data when you can.

## Assumptions and limitations

- **Detection at visits.** The event is detected at the first visit after
  it happens, and the visit schedule does not depend on the patient's risk
  beyond the observed data (non-informative inspection).
- **Biomarker visits are inspection visits.** The last visit $V$ used in
  prediction is the last biomarker measurement up to the prediction time.
- **Before the first visit.** The baseline hazard is extrapolated there, so
  the share of events before the first visit (reported by `print()`)
  indicates how much the results lean on it.
- **Stochastic EM.** Each iteration uses one draw of the event times. In an
  extreme design (visits every 2–6 years), this left a bias of about 2% of
  the effect. The pooled standard errors treat the model parameters as
  fixed, so they can be slightly too small.
- **Experimental.** The interface may change. The simulations behind the
  numbers in this article are kept in a separate folder, next to the
  package source.

## References

- Turnbull, B. W. (1976). The empirical distribution function with
  arbitrarily grouped, censored and truncated data. *Journal of the Royal
  Statistical Society, Series B*, 38, 290–295.
- Hudgens, M. G., Satten, G. A., & Longini, I. M. (2001). Nonparametric
  maximum likelihood estimation for competing risks survival data subject
  to interval censoring and truncation. *Biometrics*, 57, 74–80.
- Royston, P., & Parmar, M. K. B. (2002). Flexible parametric
  proportional-hazards and proportional-odds models for censored survival
  data. *Statistics in Medicine*, 21, 2175–2197.
- Yang, Z., Rizopoulos, D., Newcomb, L. F., & Erler, N. S. (2026).
  Time-dependent predictive accuracy metrics in the context of interval
  censoring and competing risks. *Biometrical Journal*, 68, e70108.
