---
title: "Plotting with BJM"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

BJM has plotting functions for each stage of an analysis: looking at the
data before modeling, checking the fitted sub-models, showing one patient's
predictions, and evaluating how good the predictions are. This article
walks through all of them on the `pbc3` data. Every function returns a
`ggplot` object, so you can add layers, change themes or labels with `+`,
and save it with `ggplot2::ggsave()`.

| Stage | Function | Shows |
|---|---|---|
| Data | `spaghettiPlot()` | each patient's biomarker trajectory |
| Data | `cmtPlot()` | mean trajectory of patients with the same event time |
| Data | `cifPlot()` | cumulative incidence of each event type |
| Fitted models | `plot(survival_fit_all)` | hazard and odds ratios, baseline hazard |
| Fitted models | `plot(long_fit_all)` | residual and random-effect diagnostics |
| One patient | `predictPlot()` | risk and biomarker as the horizon grows |
| One patient | `riskPlot()` | risk as the landmark time moves |
| One patient | `plot(predictLongitudinal(...))` | predicted biomarker distribution |
| Performance | `performancePlot()` | AUC and Brier score across landmarks |
| Performance | `calibrationPlot()` | predicted versus observed risk |

```{r setup, message = FALSE}
library(BJM)
data(pbc3)
```

In `pbc3`, `year` is the time of each visit, `years` the follow-up time,
`status3` is 1 for death or transplantation, and `status5` codes the event
type: 0 censored, 1 death, 2 transplantation.

## Exploring the data

### Individual trajectories: `spaghettiPlot()`

One line per patient, colored here by whether the patient died
(`status2 = 1`), with a smoothed mean trajectory per group.

```{r spaghetti, fig.alt = "Spaghetti plot of serum bilirubin against visit year, one thin line per patient, colored by whether the patient died; the smoothed mean of patients who died is higher throughout follow-up and rises over time, while that of the other patients stays near zero."}
spaghettiPlot(pbc3, bio_variable = "serBilir", time_variable = "year",
              event_type_variable = "status2")
```

With `align = "event"`, the horizontal axis becomes the time remaining
until each patient's event or censoring time, so 0 is the event and every
patient is lined up at their own end of follow-up. This is how the backward
joint model looks at a biomarker: as a function of time to the event. Here
bilirubin rises steeply in the years before death.

```{r spaghetti-event, fig.alt = "Spaghetti plot of serum bilirubin against years before each patient's event or censoring time, with the event at 0 on the left; the mean bilirubin of patients who died climbs steeply over the last few years before death, while that of the other patients rises only slightly."}
spaghettiPlot(pbc3, bio_variable = "serBilir", time_variable = "year",
              survival_variable = "years", event_type_variable = "status2",
              align = "event")
```

With many patients, `n_subjects` draws a random subset.

### Conditional mean trajectories: `cmtPlot()`

The mean biomarker trajectory of the patients whose event happened around
the same time (`condi_time2event`), here separately for the two competing
event types.

```{r cmt, message = FALSE, fig.alt = "Conditional mean trajectories of albumin for patients with an event around year 5, for the two competing event types; points are the mean albumin at each time point with a fitted line per event type, and both decline over follow-up."}
cmtPlot(pbc3[!is.na(pbc3$status4), ], condi_time2event = 5,
        event_type_variable = "status4", event_type = c("0", "1"),
        bio_variable = "albumin", time_variable = "year",
        survival_variable = "years", interval_time = 1 / 4)
```

### Cumulative incidence: `cifPlot()`

The Aalen--Johansen cumulative incidence of each event type, which treats
the competing event correctly rather than as censoring. The data may be in
long format; only each patient's first row is used.

```{r cif, fig.alt = "Cumulative incidence curves with 95% confidence bands for death and transplantation over 15 years of follow-up; death reaches about 0.6 while transplantation levels off near 0.1."}
cifPlot(pbc3, survival_variable = "years", event_type_variable = "status5",
        event_labels = c("1" = "Death", "2" = "Transplant"))
```

`group_variable` draws one curve per group, e.g. treatment arm:

```{r cif-group, fig.alt = "Cumulative incidence of death by treatment arm, placebo versus D-penicillamine, with overlapping 95% confidence bands; the two curves are very close over the whole follow-up."}
cifPlot(pbc3, survival_variable = "years", event_type_variable = "status2",
        group_variable = "drug")
```

## Checking the fitted sub-models

We fit the sub-models as in `vignette("BJM-intro")`: a Cox model for the
overall event with a logistic model for the event type, and two
biomarkers.

```{r fit}
survival_fit_all <- survivalSub(
  pbc3[!duplicated(pbc3$id), ],
  form_marginal_surv = Surv(years, status3) ~ age + sex,
  form_conditional_cr = status4 ~ years + age + sex
)

long_sub_fixed <- list(
  "serBilir" = serBilir ~ year + age + sex + years + years * year,
  "albumin"  = albumin  ~ year + age + sex + years + years * year
)
long_sub_random <- list("serBilir" = ~ year | id, "albumin" = ~ year | id)
long_fit_all <- longitudinalSub(pbc3[pbc3$status3 == 1, ], long_sub_fixed, long_sub_random)

trans <- survivalTrans(c(1, 3, 5, 7))
```

### Survival sub-model: `plot(survival_fit_all)`

A forest plot of the hazard ratios of the survival model and, with
competing risks, the odds ratios of the event-type model (here the odds of
transplantation, `status4 = 1`, versus death).

```{r plot-survival, fig.alt = "Forest plot in two panels: hazard ratios of the survival model, where age is slightly above 1 and sex is below 1 with a wide interval, and odds ratios of the event-type model for age, sex and event time, all close to 1 except age, slightly below 1."}
plot(survival_fit_all)
```

`which = "basehaz"` shows the Cox model's baseline cumulative hazard.

```{r plot-basehaz, fig.alt = "Step curve of the Cox model's baseline cumulative hazard, increasing steadily from 0 to about 0.65 over 15 years of follow-up."}
plot(survival_fit_all, which = "basehaz")
```

### Longitudinal sub-models: `plot(long_fit_all)`

Standardized residuals against fitted values, one panel per biomarker. A
trend in the red smoother, or a funnel shape, points to a misspecified mean
or non-constant variance.

```{r plot-residuals, fig.alt = "Standardized residuals against fitted values for bilirubin and albumin, one panel each; points scatter around zero and the red smoothers stay close to zero, with mild curvature at the extremes."}
plot(long_fit_all)
```

`which = "qq"` checks the residuals' normality, and `which = "ranef"` that
of the predicted random effects.

```{r plot-ranef, fig.height = 6, fig.alt = "Normal Q-Q plots of the predicted random intercepts and slopes of bilirubin and albumin, four panels; the intercepts follow the reference line closely, while the slopes have somewhat heavier tails."}
plot(long_fit_all, which = "ranef")
```

`which = "corr"` shows the correlations of the random effects across
biomarkers, estimated by the multivariate mixed model: here, patients whose
bilirubin rises faster tend to have albumin that falls faster.

```{r plot-corr, fig.width = 6, fig.height = 5, fig.alt = "Heat map of the correlations between the random intercepts and slopes of bilirubin and albumin; the strongest is -0.50, between the bilirubin slope and the albumin slope."}
plot(long_fit_all, which = "corr")
```

## Predictions for one patient

Patient 2's measurements up to year 5. In a real prediction, later
measurements and the event time are not known yet.

```{r patient}
patient <- pbc3[pbc3$id == 2 & pbc3$year <= 5, ]
patient$years <- NA
patient$status4 <- NA
```

### Risk over the prediction horizon: `predictPlot()`

The landmark is fixed at year 5 and the prediction window grows. The plot
shows the patient's bilirubin history, the predicted risk of each event
type, and the predicted future bilirubin.

```{r predict-plot, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5.5, fig.alt = "Prediction plot for patient 2 at year 5: the observed bilirubin history up to year 5, the predicted future bilirubin with widening uncertainty bands from year 5 to 8, and the predicted risk of the two event types growing with the horizon, death to about 0.25 by year 8 and transplantation staying near 0."}
predictPlot(list(patient, patient), long_fit_all, survival_fit_all,
            prediction_time = 5, horizon = seq(0, 3, 0.5), time_variable = "year",
            trans$survival_variable_all, trans$survival_trans_function,
            bandcount1 = 10, bandcount2 = 20, bandcount3 = 50)
```

### Risk as the landmark moves: `riskPlot()`

Here the window is fixed at one year and the landmark moves, so each point
uses more of the patient's history.

```{r risk-plot, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5.5, fig.alt = "Patient 2's bilirubin history from year 0 to 5 together with the predicted one-year risks of the two event types at landmarks 1 to 5; the risk of death rises slowly as bilirubin increases, and the transplantation risk stays near 0."}
riskPlot(list(patient, patient), long_fit_all, survival_fit_all,
         prediction_time = c(1, 2, 3, 4, 5), bio_i = 1, horizon = 1,
         time_variable = "year", trans$survival_variable_all,
         trans$survival_trans_function, bandcount1 = 10, bandcount2 = 20)
```

### Predicted biomarker distribution: `plot(predictLongitudinal(...))`

The predicted distribution of bilirubin one year after the landmark, for
three patients, with the most likely value marked by a dashed line.

```{r predict-longitudinal, warning = FALSE, fig.alt = "Predicted density curves of bilirubin one year after year 3 for patients 2, 4 and 5, each with a dashed line at its most likely value; the curves have similar spread and are centred at increasing values, patient 2 lowest and patient 5 highest."}
three <- pbc3[pbc3$id %in% c(2, 4, 5) & pbc3$year <= 3, ]
three$years <- NA
three$status4 <- NA
pred <- predictLongitudinal(list(three, three), long_fit_all, survival_fit_all,
                            prediction_time = 3, horizon = 1, time_variable = "year",
                            trans$survival_variable_all, trans$survival_trans_function,
                            bio_i = 1, bandcount2 = 20, bandcount3 = 50)
plot(pred)
```

## Evaluating the predictions

Performance measured on the data the model was fit on is optimistic. Here
we split the patients at random, fit the sub-models on two thirds of them,
and evaluate on the remaining third.

```{r split}
set.seed(2026)
ids <- unique(pbc3$id)
train_ids <- sample(ids, round(2 / 3 * length(ids)))
train <- pbc3[pbc3$id %in% train_ids, ]
test <- pbc3[!pbc3$id %in% train_ids, ]

survival_train <- survivalSub(train[!duplicated(train$id), ],
                              Surv(years, status3) ~ age + sex,
                              status4 ~ years + age + sex)
long_train <- longitudinalSub(train[train$status3 == 1, ], long_sub_fixed, long_sub_random)
```

### Discrimination and accuracy: `performancePlot()`

At each landmark, every test patient still event-free is given a predicted
risk of an event in the next two years, from their history so far. The
time-dependent AUC measures how well these risks separate the patients who
have the event from those who do not (0.5 is chance, 1 is perfect); the
Brier score is the mean squared error of the risks (lower is better).
Censoring is handled by inverse probability of censoring weighting.

```{r performance, warning = FALSE, fig.alt = "Time-dependent AUC and Brier score at landmarks 1 to 5 for death and transplantation on the test patients; AUCs are well above the chance line of 0.5 and Brier scores stay below 0.1, with the transplantation AUC missing at landmark 1."}
performancePlot(test, long_train, survival_train,
                prediction_time = c(1, 2, 3, 4, 5), horizon = 2, time_variable = "year",
                trans$survival_variable_all, trans$survival_trans_function,
                bandcount1 = 20, bandcount2 = 80)
```

No test patient had a transplant in the two years after year 1, so the
transplant AUC is undefined there and that point is left out (the message
above). The returned plot's `data` holds the numbers, including how many
patients were at risk and how many had each event at each landmark.

### Calibration: `calibrationPlot()`

The AUC only checks the ranking of the risks. Calibration checks the
values: the test patients are grouped by predicted risk, and each group's
mean predicted risk is plotted against the risk actually observed in it.
Points near the diagonal mean the predictions are accurate; below it, the
model overestimates the risk, above it, it underestimates it.

```{r calibration, warning = FALSE, fig.height = 6, fig.alt = "Calibration plots for landmarks 2 and 4 and the two event types: mean predicted against observed risk in four risk groups with 95% confidence intervals; the points lie near the diagonal, with wide intervals where few events occurred."}
calibrationPlot(test, long_train, survival_train,
                prediction_time = c(2, 4), horizon = 2, time_variable = "year",
                trans$survival_variable_all, trans$survival_trans_function,
                n_groups = 4, bandcount1 = 20, bandcount2 = 80)
```

With only about a hundred test patients, and few transplants among them,
the observed risks have wide confidence intervals; use more groups with
larger data.
